---
title: "Congress Data Package"
author: "JTD"
format: html
editor: visual
---


```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggthemes)

theme_set(theme_minimal(base_size = 12))
```

## Raw Data

the raw data was obtained from 
[this site](https://congressdata.joshuamccrain.com/).
It was published in 2016 by Ella Foster-Molina, affiliated with the University of Rochester
It contains information about each congressional district from years 1972-2013, as well as personal characteristics of the house member representing that district.
This dataset was likely created for use in research papers.

More information about the dataset can be found [here](codebook.pdf).

let's load the data:
```{r}
congress_data_raw <- read_csv("data/allCongressDataPublishV2.csv")
```

```{r}
glimpse(congress_data_raw)
```

```{r}
congress_data %>%
  ggplot(aes(x=congNum, group=prcntWhite, fill=prcntWhite,)) +
  geom_bar()
```

```{r}
congress_data_raw %>%
  filter(party=="Republican-Conservative")
```
```{r}
congress_data_raw %>%
  filter(is.na(congNum))
```

```{r}
#save this as a new table!
congress_data_raw %>%
  group_by(congNum, partyControl, demSeats, repSeats, otherSeats) %>%
  summarise()
```
```{r}
congress_data%>%
  filter(sponID==406746) %>%
  select(party)

```

```{r}
congress_data %>%
  filter(!is.na(prcntBlackNotHisp)) %>%
  select(congNum) %>%
  distinct()
```

```{r}
congress_data %>%
  select(prcntWhite)
```


