---
title: "Congress Data Modeling"
author: "JTD"
format: html
---
```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
#get testing data
set.seed(123)

#load data
congress_data <- read_csv("data/congress_data.csv")

congress_data <- congress_data %>%
  mutate(party_binary = ifelse(party=="Democrat", 1, 0),
         gender_binary = ifelse(gender=="M", 1, 0))

congress_data_split <-
  initial_split(congress_data, prop = 0.8, strata = party)

congress_data_train <- training(congress_data_split)

congress_data_test <- testing(congress_data_split)

#get specification
gam_spec <- 
  gen_additive_mod() %>%
  set_mode("regression")

#get fit for # of bills sponsored
numspon_fit <- gam_spec %>%
  fit(numSpon ~ s(numberTerms) + s(age) + gender_binary + party_binary, data=congress_data_train)

#predict using the model
numspon_preds <- numspon_fit %>%
  predict(congress_data_test) %>%
  cbind(congress_data_test)


#get fit for pass percent
passprcnt_fit <- gam_spec %>%
  fit(passPrcnt ~ s(numberTerms) + s(age) + gender_binary + party_binary, data=congress_data_train)

#predict using the model
passprcnt_preds <- passprcnt_fit %>%
  predict(congress_data_test) %>%
  cbind(congress_data_test)
```
The script used for the following models can be found in [R/models.R](R/models.R).

These models serve to predict both the number of bills passed by each representative, as well as the pass percentage of their bills introduced. 

The predictors for each model are the age, gender, number of terms served, and party of each representative.

Let's take a look at the RMSE values for each model, and compare the predicted values to the actual values.

***

```{r}
#| echo: false
#find the RMSE
numspon_preds %>% 
  rmse(truth = numSpon, estimate = .pred)

#compare predictions to actual
numspon_preds %>%
  select(numSpon, .pred) %>%
  glimpse()
```
***

It seems that the first model is not quite accurate. Since the RMSE value is 17.3, this means that on average, the model prediction for number of bills sponsored is off by 17.3 bills.

***
```{r}
#| echo: false
#find the RMSE
passprcnt_preds %>% 
  rmse(truth = passPrcnt, estimate = .pred)

#compare predictions to actual
passprcnt_preds %>%
  select(passPrcnt, .pred) %>%
  glimpse()
```
***

Similarly, the model for pass percentage also has a large margin of error at 10.9 percent.


It seems that based on these models, these variables do not serve very well as predictors for the number of bills sponsored and pass percentage. From this model analysis, one could conclude that personal characteristics do not have a considerable effect on a person's effectiveness in congress.
