---
title: "Congress Data Analysis"
subtitle: "Impact of Personal Characteristics on Congressional Effectiveness"
author: "Joseph Dolan"
format: revealjs
bibliography: references.bib
---
```{r}
#| echo: false
#| message: false
#| warning: false

#load libraries
library(tidyverse)
library(ggthemes)
library(patchwork)
library(ggplot2)
library(ISLR2)
library(rsample)
library(tidymodels)
library(kableExtra)

#load data
source("../R/clean_congress_data.R")
set.seed(123)

theme_set(theme_minimal(base_size = 12))
congress_data <- read_csv("../data/congress_data.csv")
```

## Initial Research Question

- Congress is made up of hundreds of representatives, all of who have different backgrounds.

- *Do these personal characteristics have an effect on the number of bills they sponsor, or the percentage of bills they pass?*

## Congressional Dataset

```{r}
#| echo: false
congress_data %>%
  head(5) %>%
  kable()
```
[@congress_dataset]

## Party Distribution of Congress Over Time
```{r}
#| echo: false
#| warning: false
congress_data %>%
  ggplot(aes(x=congNum, fill=party)) +
  geom_bar() +
  scale_fill_manual(values=c("#5156e0", "#00aa00", "#e05151")) +
  labs(
    x="Congress Number",
    y="Number of Seats",
    fill="Party"
       )
```

## Gender Distribution of Congress
```{r}
#| echo: false
#| warning: false
congress_data %>%
  ggplot(aes(x=congNum, fill=gender)) +
  geom_bar(color="black") +
  scale_fill_manual(values=c("#e05151", "#5156e0")) +
  labs(
    x="Congress Number",
    y="Number of Seats",
    fill="Gender"
       )
```

## Age of Congress Over Time

```{r}
#| echo: false
#| warning: false
congress_data %>%
  ggplot(aes(x=congNum, y=age)) +
  geom_point(color="lightblue", alpha=0.5) +
  geom_smooth(color="#512d6d",fill="darkgreen") +
  labs(x="Congress Number",y="Average Age")
```

## Number of Bills Sponsored v. Number of Terms Served
```{r}
#| echo: false
#| warning: false
congress_data %>%
  group_by(numberTerms=numberTerms%/%3) %>%
  summarise(mean_numSpon = mean(numSpon)) %>%
  ggplot(aes(x=factor(numberTerms*3), y=mean_numSpon)) +
  geom_bar(stat='identity', fill="#553E4E") +
  labs(
    x="Number of Terms Served",
    y="Average # of Bills Sponsored"
  )

```

## Pass Percentage v. Number of Terms Served

```{r}
#| echo: false
#| warning: false

congress_data %>%
  filter(!is.na(passPrcnt)) %>%
  group_by(numberTerms=numberTerms%/%3) %>%
  summarise(mean_passprcnt = mean(passPrcnt)) %>%
  ggplot(aes(x=factor(numberTerms*3), y=mean_passprcnt)) +
  geom_bar(stat='identity', fill="#553E4E") +
  labs(
    x="Number of Terms Served",
    y="Pass Percentage"
  )

```

## Number of Bills Sponsored v. Age
```{r}
#| echo: false
#| warning: false

congress_data %>%
  group_by(age=age%/%5) %>%
  summarise(mean_numSpon = mean(numSpon)) %>%
  ggplot(aes(x=factor(age*5), y=mean_numSpon)) +
  geom_bar(stat='identity', fill="#553E4E") +
  labs(x="Age Group", y="Average # of Bills Sponsored")
```

## Pass Percentage v. Age
```{r}
#| echo: false
#| warning: false

congress_data %>%
  filter(!is.na(passPrcnt)) %>%
  group_by(age=age%/%5) %>%
  summarise(mean_passprcnt = mean(passPrcnt)) %>%
  ggplot(aes(x=factor(age*5), y=mean_passprcnt)) +
  geom_bar(stat='identity', fill="#553E4E") +
  labs(
    x="Age Group",
    y="Pass Percentage"
  )

```
## Number of Bills Sponsored v. Gender
```{r}
#| echo: false
#| warning: false
congress_data %>%
  group_by(gender) %>%
  ggplot(aes(x=gender, y=numSpon, color=gender)) +
  scale_color_manual(values=c("#e05151", "#5156e0")) +
  geom_boxplot(outlier.shape = NA, color="black") +
  geom_jitter(width=0.35, alpha=0.2) +
  ylim(0, 50) +
  labs(x="", y="# of Bills Sponsored", color="Gender")
```
## Pass Percentage v. Gender
```{r}
#| echo: false
#| warning: false

congress_data %>%
  filter(passPrcnt>0) %>%
  group_by(gender) %>%
  ggplot(aes(x=gender, y=passPrcnt, color=gender)) +
  scale_color_manual(values=c("#e05151", "#5156e0")) +
  geom_boxplot(outlier.shape = NA, color="black") +
  geom_jitter(width=0.35, alpha=0.2) +
  ylim(0, 50) +
  labs(x="", y="Pass Percentage", color="Gender")
```

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
#get testing data
set.seed(123)

#load data
congress_data <- read_csv("../data/congress_data.csv")

congress_data <- congress_data %>%
  mutate(party_binary = ifelse(party=="Democrat", 1, 0),
         gender_binary = ifelse(gender=="M", 1, 0))

congress_data_split <-
  initial_split(congress_data, prop = 0.8, strata = party)

congress_data_train <- training(congress_data_split)

congress_data_test <- testing(congress_data_split)

#get specification
gam_spec <- 
  gen_additive_mod() %>%
  set_mode("regression")

#get fit for # of bills sponsored
numspon_fit <- gam_spec %>%
  fit(numSpon ~ s(numberTerms) + s(age) + gender_binary + party_binary, data=congress_data_train)

#predict using the model
numspon_preds <- numspon_fit %>%
  predict(congress_data_test) %>%
  cbind(congress_data_test)


#get fit for pass percent
passprcnt_fit <- gam_spec %>%
  fit(passPrcnt ~ s(numberTerms) + s(age) + gender_binary + party_binary, data=congress_data_train)

#predict using the model
passprcnt_preds <- passprcnt_fit %>%
  predict(congress_data_test) %>%
  cbind(congress_data_test)
```

## Model Predictions for Number of Bills Sponsored
- RMSE: 17.3
```{r}
#| echo: false
#| message: false

#compare predictions to actual
numspon_preds %>%
  select(numSpon, .pred)%>%
  top_n(4)%>%
  kbl(caption = "# of Bills Sponsored: Predictions vs Actual")
```

## Model Predictions for Pass Percentage
- RMSE: 10.9
```{r}
#| echo: false
#| message: false

#compare predictions to actual
passprcnt_preds %>%
  select(passPrcnt, .pred)%>%
  top_n(4)%>%
  kbl(caption = "Pass Percentage: Predictions vs Actual")
```

## Bootstrap Estimates for Number of Bills Sponsored
```{r}
#| echo: false
#| message: false
# clean up data for ANOVA
congress_data_for_anova <-congress_data %>%
  filter(party!="Independent")

#-----numSpon-----

#model formula
numSpon_lm <-
  lm(numSpon ~ numberTerms + age + gender + party, data = congress_data_for_anova)

#get intervals
numSpon_intervals <- 
  reg_intervals(numSpon ~ numberTerms + age + gender + party, data = congress_data_for_anova, times = 500)

#plot estimates
numSpon_intervals %>%
  ggplot(aes(.estimate, term)) +
  geom_vline(xintercept = 0, linewidth = 1.5, lty = 2, color = "gray50") +
  geom_errorbar(linewidth = 1.4, alpha = 0.7,
                aes(xmin = .lower, xmax = .upper)) +
  geom_point(size = 3)
```

## Bootstrap Estimates for Pass Percentage
```{r}
#| echo: false
#| message: false
#-----passPrcnt-----

#model formula
passPrcnt_lm <-
  lm(passPrcnt ~ numberTerms + age + gender + party, data = congress_data_for_anova)

#get intervals
passPrcnt_intervals <- 
  reg_intervals(passPrcnt ~ numberTerms + age + gender + party, data = congress_data_for_anova, times = 500)

#plot estimates
passPrcnt_intervals %>%
  ggplot(aes(.estimate, term)) +
  geom_vline(xintercept = 0, linewidth = 1.5, lty = 2, color = "gray50") +
  geom_errorbar(linewidth = 1.4, alpha = 0.7,
                aes(xmin = .lower, xmax = .upper)) +
  geom_point(size = 3)
```

## Conclusions

- There is a lack of evidence that personal characteristics affect effectiveness in congress.

- Further research could investigate other possible predictors, such as demographics of each district.

## References